---
title: "Dimensionality reduction of the Cooney (C057) memory CD4+ T-cell data set"
description: |
author:
  - name: Peter Hickey 
    url: https://peterhickey.org
    affiliation: Single Cell Open Research Endeavour (SCORE), WEHI
    affiliation_url: https://www.wehi.edu.au/people/shalin-naik/3310/score
date: "`r Sys.Date()`"
output: radix::radix_article
editor_options: 
  chunk_output_type: console
bibliography: ref.bib
---

```{r setup}
library(SingleCellExperiment)
library(here)
library(rmarkdown)
library(BiocStyle)
library(scran)
library(janitor)
library(ggplot2)
library(cowplot)
source(here("analysis", "helper_functions.R"))
# NOTE: Using >= 4 cores siezes up my laptop. Can use more on RStudio server.
options(
  "mc.cores" = 
    ifelse(Sys.info()[["nodename"]] == "rstudio-2.hpc.wehi.edu.au", 8L, 2L))
register(MulticoreParam(workers = getOption("mc.cores")))
knitr::opts_chunk$set(fig.path = "C057_Cooney_merge_files/")
```

# Setting up the data

We start from the preprocessed *SingleCellExperiment* object created in ['Preprocessing the Cooney (C057) memory CD4+ T-cell data set'](C057_Cooney_preprocess.html).

```{r}
sce <- readRDS(here("data", "SCEs", "C057_Cooney.preprocessed.SCE.rds"))
```

```{r}
hto_to_sample_df <- as.data.frame(colData(sce)) %>%
  dplyr::select(hto_cluster, HTO, Sample, Treatment, Replicate) %>%
  dplyr::distinct()

# Some useful colours
hto_cluster_colours <- setNames(
  colorblindr::palette_OkabeIto,
  levels(sce$hto_cluster))
hto_cluster_colours <- hto_cluster_colours[!is.na(names(hto_cluster_colours))]
hto_colours <- hto_cluster_colours
names(hto_colours) <- dplyr::inner_join(
  data.frame(hto_cluster = names(hto_cluster_colours)),
  hto_to_sample_df) %>%
  dplyr::pull(HTO)
sample_colours <- hto_cluster_colours
names(sample_colours) <- dplyr::inner_join(
  data.frame(hto_cluster = names(sample_colours)),
  hto_to_sample_df) %>%
  dplyr::pull(Sample)
treatment_colours <- c(
  "Infected" = "#e41a1c",
  "Uninfected" = "#377eb8",
  "Unknown" = "grey70")
# NOTE: No replicate_colours because we won't ever colour by this.
```

# Feature selection

## Motivation

We often use scRNA-seq data in exploratory analyses to characterize heterogeneity across cells.
Procedures like clustering and dimensionality reduction compare cells based on their gene expression profiles, which involves aggregating per-gene differences into a single (dis)similarity metric between a pair of cells.
The choice of genes to use in this calculation has a major impact on the behavior of the metric and the performance of downstream methods.
We want to select genes that contain useful information about the biology of the system while removing genes that contain random noise.
This aims to preserve interesting biological structure without the variance that obscures that structure.
It also reduces the size of the dataset to improve computational efficiency of later steps.

## Quantifying per-gene variation

The simplest approach to feature selection is to select the most variable genes based on their expression across the population.
This assumes that genuine biological differences will manifest as increased variation in the affected genes, compared to other genes that are only affected by technical noise or a baseline level of 'uninteresting' biological variation (e.g., from transcriptional bursting).

Key to this approach is estimating the 'uninteresting' variation.
For droplet-based scRNA-seq, like 10X Genomics, we do not have spike-in RNAs that we could use to estimate the technical variation.
In the absence of spike-in data, one can attempt to create a trend by making some distributional assumptions about the noise.
For example, UMI counts typically exhibit near-Poisson variation if we only consider technical noise from library preparation and sequencing.
This is the approach we use in the below.

```{r}
# Fit a mean-variance trend to the endogenous genes.
var_fit <- trendVar(
  sce,
  parametric = FALSE,
  loess.args = list(span = 0.05),
  use.spikes = FALSE)
# Estimate a mean-variance under the assumption of Poisson variation.
var_fit$trend <- makeTechTrend(x = sce)
```

Figure \@ref(fig:mean-var-plot) shows the mean-variance relationship for each gene in the dataset and the estimated trend.
Interestingly, trends based purely on technical noise tend to yield large biological components for highly-expressed genes. 
This often includes so-called "house-keeping" genes coding for essential cellular components such as ribosomal proteins, which are often considered uninteresting for characterizing cellular heterogeneity.
However, one should keep an open mind with respect to the 'interesting-ness' of house-keeping genes because these are regularly differentially expressed in a variety of conditions and the fact that they have large biological components indicates that there is strong variation across cells that may not be completely irrelevant.

```{r mean-var-plot, fig.cap = "Variance of normalized log-expression values for each gene in the dataset, plotted against the mean log-expression. The blue line represents represents the mean-variance relationship corresponding to Poisson noise."}
plot(var_fit$means, var_fit$vars)
curve(var_fit$trend(x), col ="dodgerBlue", add = TRUE, lwd = 2)
```

We decompose the variance for each gene using the Poisson-based trend, and examine the genes with the highest biological components.
This ensures that the variance estimate is not driven by one or two outlier cell.
Figure \@ref(fig:hvgvioplot) shows this for the combined dataset.

```{r hvgvioplot, fig.cap = "Violin plots of normalized log-expression values for the top 10 genes with the largest biological components in the combined dataset. Each point represents the log-expression value in a single cell."}
var_out <- decomposeVar(sce, var_fit)
chosen_genes <- order(var_out$bio, decreasing = TRUE)[1:10]

library(scater)
plotExpression(
  object = sce, 
  features = rownames(var_out)[chosen_genes])
```

## Selecting highly variable genes

We define the features of interest as those with net biological components greater than zero.

```{r}
to_use <- var_out$bio > 0
```

This enriches for genes that contain some biological variation, reducing the effect of uninteresting Poisson noise on downstream analyses.
There are `r sum(to_use)` such genes.

# Dimensionality reduction

## Overview

Many scRNA-seq analysis procedures involve comparing cells based on their expression values across multiple genes.
For example, clustering aims to identify cells with similar transcriptomic profiles by computing 'distances' across genes.
In these applications, each individual gene represents a dimension of the data.
More intuitively, if we had a scRNA-seq data set with two genes, we could make a two-dimensional plot where each axis represents the expression of one gene and each point in the plot represents a cell.
This concept can be extended to data sets with thousands of genes where each cell’s expression profile defines its location in the high-dimensional expression space.

As the name suggests, dimensionality reduction aims to reduce the number of separate dimensions in the data.
This is possible because different genes are correlated if they are affected by the same biological process.
Thus, we do not need to store separate information for individual genes, but can instead compress multiple features into a single dimension.
This has several benefits:

- It reduces computational work in downstream analyses, as calculations only need to be performed for a few dimensions rather than thousands of genes
- It reduces noise by averaging across multiple genes to obtain a more precise representation of the patterns in the data
- It enables effective plotting of the data

## Principal components analysis

Principal components analysis (PCA) discovers axes in high-dimensional space that capture the largest amount of variation. 
This is best understood by imagining each axis as a line.
Say we draw a line anywhere, and we move all cells in our data set onto this line by the shortest path.
The variance captured by this axis is defined as the variance across cells along that line.
In PCA, the first axis (or "principal component", PC) is chosen such that it captures the greatest variance across cells.
The next PC is chosen such that it is orthogonal to the first and captures the greatest remaining amount of variation, and so on.

By definition, the top PCs capture the dominant factors of heterogeneity in the data set.
Thus, we can perform dimensionality reduction by restricting downstream analyses to the top PCs.
This strategy is simple, highly effective and widely used throughout the data sciences.
It takes advantage of the well-studied theoretical properties of the PCA - namely, that a low-rank approximation formed from the top PCs is the optimal approximation of the original data for a given matrix rank.
It also allows us to use a wide range of fast PCA implementations for scalable and efficient data analysis.

When applying PCA to scRNA-seq data, our assumption is that biological processes affect multiple genes in a coordinated manner.
This means that the earlier PCs are likely to represent biological structure as more variation can be captured by considering the correlated behaviour of many genes.
By comparison, random technical or biological noise is expected to affect each gene independently.
There is unlikely to be an axis that can capture random variation across many genes, meaning that noise should mostly be concentrated in the later PCs.
This motivates the use of the earlier PCs in our downstream analyses, which concentrates the biological signal to simultaneously reduce computational work and remove noise.

We perform the PCA on the log-normalized expression values.

```{r}
set.seed(861)
sce <- denoisePCA(
  sce, 
  technical = var_out, 
  BSPARAM = BiocSingular::IrlbaParam(deferred = TRUE))
```

## Dimensionality reduction for visualization

Another application of dimensionality reduction is to compress the data into 2 (sometimes 3) dimensions for plotting.
This serves a separate purpose to the PCA-based dimensionality reduction described above.

### $t$-stochastic neighbor embedding

The de facto standard for visualization of scRNA-seq data is the $t$-stochastic neighbour embedding [@van2008visualizing]. 
This attempts to find a low-dimensional representation of the data that preserves the distances between each point and its neighbours in the high-dimensional space.
Unlike PCA, it is not restricted to linear transformations, nor is it obliged to accurately represent distances between distance populations.
This means that it has much more freedom in how it arranges cells in low-dimensional space, enabling it to separate many distinct clusters in a complex population.
Figure \@ref(fig:tsne) shows the $t$-SNE plot for the dataset, coloured by `Sample.

```{r tsne, fig.cap = "t-SNE plot constructed from the top PCs in the dataset. Each point represents a cell, coloured according to the `Sample`.", fig.asp = 0.8}
set.seed(811771)
sce <- runTSNE(sce, use_dimred = "PCA")
tsne_df <- cbind(
  data.frame(
    x = reducedDim(sce, "TSNE")[, 1],
    y = reducedDim(sce, "TSNE")[, 2]),
  as.data.frame(colData(sce)))

ggplot(aes(x = x, y = y), data = tsne_df) +
  geom_point(aes(colour = Sample), size = 0.5) +
  scale_colour_manual(values = sample_colours) +
  theme_cowplot(font_size = 10) + 
  ggtitle("t-SNE") +
  xlab("Dimension 1") + 
  ylab("Dimension 2")
```

### Interpreting the plots

We can start to explore what biological factors might be $t$-SNE results.
Figure \@ref(fig:tsne-more-factors) shows a strong separation of `Unknown` samples from the `Treated` and `Untreated` samples, and a secondary separation between many of the `Treated` and `Untreated` sampels.

```{r tsne-more-factors, fig.cap = "t-SNE plot constructed from the top PCs in the dataset. Each point represents a cell, coloured to highlight the effect of various factors.", fig.asp = 1.5, layout = "l-body"}
bg <- dplyr::select(tsne_df, -Treatment, -Sample)
ggplot(aes(x = x, y = y), data = tsne_df) +
  geom_point(data = bg, colour = scales::alpha("grey", 0.5), size = 0.125) +
  geom_point(aes(colour = Sample), alpha = 1, size = 0.25) +
  scale_fill_manual(values = sample_colours, name = "Sample") + 
  scale_colour_manual(values = sample_colours, name = "Sample") +
  theme_cowplot() + 
  xlab("Dimension 1") + 
  ylab("Dimension 2") +
  facet_grid(Sample ~ Treatment)
```

Dimensionality reduction for visualization necessarily involves discarding information and distorting the distances between cells in order to fit high-dimensional data into a 2-dimensional space.
One might wonder whether the results of such extreme data compression can be trusted.

As a general rule, we would not perform clustering on the $t$-SNE coordinates.
Rather, we would cluster on the first 10-50 PCs and then visualize the cluster identities on the $t$-SNE plot.
This ensures that clustering makes use of the information that was lost during compression into two dimensions.
The $t$-SNE plot can then be used for a diagnostic inspection of the clustering output.
In particular, the plot is most useful for checking whether two clusters are actually neighboring subclusters or whether a cluster can be split into further subclusters, which are generally safe interpretations of $t$-SNE coordinates.

It is worth elaborating on why we should not perform downstream analyses directly on the $t$-SNE coordinates.
Let us put aside the fact that operating on the high-dimensional representations preserves more information; from a naive perspective, using the $t$-SNE coordinates is very tempting as it ensures that any results are immediately consistent with the visualizations (regardless of whether they are right).
However, this can actually be considered a disservice as it masks the uncertainty of the results, leading us to place more confidence in them than is warranted.
Rather than being errors, major discrepancies can instead be useful for motivating further investigation into the more ambiguous parts of the dataset; conversely, the lack of discrepancies increases trust in the results.

# Concluding remarks

```{r}
saveRDS(
  sce,
  here("data", "SCEs", "C057_Cooney.merged.SCE.rds"),
  compress = "xz")
```

The processed *SingleCellExperiment* object is available (see [`data/SCEs/C057_Cooney.merged.SCE.rds`](../data/SCEs/C057_Cooney.merged.SCE.rds)).
This will be used in downstream analyses, e.g., annotating cell types, differential expression analysis between conditions within each cluster, and differential abundance analyses between conditions for each cluster.

# Additional information {.appendix}

The following are available on request:

- Full CSV tables of any data presented.
- PDF/PNG files of any static plots.

# Session info {.appendix}

The analysis and this document were prepared using the following software:

```{r}
sessionInfo()
```
